Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-30 2024pipit/section/20-Motivation.tex
doublegood

\autoref{f:tt-systemmatrix-ok} (left) shows an example node matrix for a non-leader node.
The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering a failure if the reference message is not received.
\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.

@@

Our driver may fall behind when trying to run certain schedules, as the driver only processes one trigger per time-mark.
To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.

Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
However, we could \emph{not} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.

We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.

@@

The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous tick of computation.
Expressing such a rely in an imperative setting would require extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.

>>>>

\autoref{f:tt-systemmatrix-ok} (left) shows an example node matrix for a non-leader node.
The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering an error if the reference message is not received.
\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.

@@

Our driver may fall behind when trying to execute certain schedules, as the driver only processes one trigger per time-mark.
To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.

Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
However, we \emph{cannot} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.

We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index, so that it is reachable in time from the start of the array.

@@

The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous \emph{tick} of computation.
Expressing such a rely in an imperative setting would require extra encoding, as preconditions in imperative languages do not generally have an innate notion of the previous value with respect to a global shared clock.

### 2024-03-30 2024pipit/section/30-Core.tex
doublegood

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.

@@

At a high level, evaluation unfolds recursive streams to determine a value.
For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumionetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
Evaluation of primitives is standard.
To evaluate variables, we look for the variable in the current heap:
$$
\ruleIN{
}{
  \bigstep{\sumionetwo}{\textit{ints}}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumionetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

@@

Checking contracts is more involved.
Conceptually, contracts involve two kinds of checks: one for the definition and one for the call-site.
To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
% The fact that the checked semantics refers to a particular $\Sigma$ is significant: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.

@@

It is critical that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
The contract itself remains valid, as the input did not satisfy the rely at all steps.
This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.

>>>>

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.
% The unknown property status here allows us to defer the caller's proof that the input stream is always positive until we have a concrete definition for the \emph{ints} input stream.

@@

At a high level, evaluation unfolds recursive streams to determine a value.
For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumintsonetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
Evaluation of primitives is standard.
To evaluate variables, we look for the variable in the current (rightmost) heap:
$$
\ruleIN{
}{
  \bigstep{\sumintsonetwo}{\textit{ints}}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumintsonetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

@@

Checking contracts is more involved.
For whole-program correctness, it would suffice to check that a contract's rely and guarantee both hold.
However, the purpose of contracts is to enable modular reasoning about parts of the program: we need to be able to check contracts independently of their context.
Conceptually, then, contracts involve two kinds of checks: one for the definition and one for the call-site.
To check a contract definition, we check that the body satisfies the guarantee for all \emph{valid} contexts -- that is, those where the rely holds.
Then, to check a contract instance, we just need to check that the call-site satisfies the rely.
% To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we want to check that the body satisfies the guarantee
% check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
% The fact that the checked semantics refers to a particular $\Sigma$ is significant: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.

@@

It is critical that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
The contract itself remains valid, however, as the assumption is invalid: the input did not satisfy the rely at all steps.
% The \emph{at all steps} is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

% Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.

### 2024-03-30 2024pipit/section/50-Extraction.tex
doublegood

To execute recursive expressions $\xrec{x}{e} : \tau$, we require an arbitrary value of type $\tau$ to seed the fixpoint, as described in \autoref{s:core:causality}.
We first call the step function to evaluate $e$ with $x$ bound to $\bot_\tau$.
This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
To get the correct state, we call the step function again, this time with $e$ bound to $v$.

@@

Here, the step function takes heaps of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value used in the state.
This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.

>>>>

To execute recursive expressions $\xrec{x}{e} : \tau$, we require an arbitrary value of type $\tau$ to seed the fixpoint, as described in \autoref{s:core:causality}.
We first call the step function to evaluate $e$ with $x$ bound to $\bot_\tau$.
This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
To get the correct state, we call the step function again, this time with $x$ bound to the correct value, $v$.

@@

Here, the step function takes heaps of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value $\textit{sum}_0$ used in the state.
This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.
